### 2022-06-27 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@79b20b62fd5d3b81ec57b105439336f545a184b0 🚀

@@

Bats, the second most diverse groups of mammals, are found worldwide;
gradients in their species richness generally track broader patterns of
mammal diversity, with a striking Neotropical hotspot (especially in the
Amazon basin) and a secondary hotspot centered in the southeast Asian
peninsula. These hotspots of bat diversity are generally presumed to be
hotspots of viral adaptive radiation, and therefore areas of concern for
human health.\textsuperscript{2,13} However, the hotspots of bat
betacoronavirus reservoirs show a distinct pattern, with primary
hotspots (both in terms of size and higher values) of host richness
situated primarily southeast Asia, parts of southern Europe, and to a
lesser extent parts of Africa in the -25-0 range of latitudes
(fig.~\ref{fig:richness}; top). Although hundreds of species likely host
undiscovered betacoronaviruses, machine learning predictions have
suggested that these undiscovered reservoirs should follow the same
diversity gradient.\textsuperscript{14} In principle, these hotspots of
locally-diverse, virus-rich bat communities should drive more adaptive
diversification in their viruses.

@@

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South
America has the most evolutionary distinct hosts \emph{and} viruses,
followed by secondary hotspots in southeast Asia and the Rift Valley
region have mostly distinct viruses. Some degree of sampling bias may
contribute to these patterns: for example, South-America is one of the
places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{2,13,15} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{16}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there is
obvious heterogeneity in betacoronavirus evolutionary distinctiveness
that is distinct from overall bat diversity.

Overall, these patterns recapitulate the evolutionary history of both
the order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe
bats (Rhinolophidae) are both the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}), the group of pandemic threats that have
been of the greatest interest to researchers\textsuperscript{17} (and so
have been sampled most intensively).\textsuperscript{16} The hotspots of
host richness and viral diversity in southeast Asia---both of which are
disproportionately high, considering the global landscape of bat species
richness---are almost entirely driven by viral adaptive radiation
through host switching within this clade\textsuperscript{3,14}. In
contrast, the Neotropical hotspot of viral distinctiveness is driven by
isolation by host vicariance. Out of the four main groups of
betacoronaviruses, only the subgenus \emph{Merbecovirus} (MERS-like
viruses) has been found in animals in the Americas---an introduction
that is generally presumed to be ancient.\textsuperscript{3} While
comparatively understudied, New World merbecoviruses have been found in
the ghost-faced bats (Mormoopidae), New World leaf-nosed bats
(Phyllostomidae), and free-tailed bats (Molossidae) (add cite: Olival
2020 PLoS Pathogens). The former two groups are endemic to the
Neotropics, while the explosive adaptive radiations of the latter two
(and particularly the phyllostomids) are responsible for the hotspot of
bat diversity in the Amazon. Together, these clades of New World bats
play host to a distinct regime of betacoronavirus coevolution.

\hypertarget{global-biogeographic-regions-are-consistent-for-bats-and-betacoronaviruses}{%
\subsection{Global biogeographic regions are consistent for bats and
betacoronaviruses}\label{global-biogeographic-regions-are-consistent-for-bats-and-betacoronaviruses}}

Most previous work has assumed that coronavirus biogeography is driven
by coevolutionary regimes that form at finer taxonomic scales, treating
the presence or richness of key bat host groups as predictive of these
viruses' distribution.\textsuperscript{2,3} Projecting bat and
betacoronavirus phylogeny over space (fig.~\ref{fig:biogeo}), we find
further support for the idea that bat community assembly is directly
responsible for spatial heterogeneity in viral coevolutionary regimes.
The distinct groupings (represented by different colors, symbolizing
positions in a subspace formed by the first two phylogenetic principal
components) are essentially equivalent between the two groups, and can
be coarsely delineated as (1) south and southeast Asia, (2) east Asia
(including northern China), west Asia, and the Mediterranean coast; (3)
Eurasia above a northing of 40; and (4) Africa and south America. In
some cases, this diverges from expectations about coronavirus
biogeography: for example, previous work has rarely flagged India as a
region of interest, but for both bats and betacoronaviruses, the
subcontinent falls into the same regions as the southeast Asian
peninsula (and indeed, the region is home to known bat hosts of
nobecoviruses, sarbecoviruses, and merbecoviruses).\textsuperscript{3}

Overall, these results suggest that, although there are unique hotspots
of viral adaptive radiation, biogeographic regions are mostly consistent
between bats and their viruses. This may be surprising, given that
cospeciation plays a minor role in coronavirus
diversification,\textsuperscript{2} which could theoretically allow for
broad divergence in their biogeography. However, host jumps at the
family level or higher are relatively rare and significant events in
coronavirus evolutionary history;\textsuperscript{2,17} as a result, the
mosaic of betacoronavirus phylogeography is assembled from a set of
overlapping smaller coevolutionary systems, superimposed in space and
filtered by the importance of different subgroups in local host
communities. For example, the most speciose and cosmopolitan family of
bats, the vesper bats (Vespertilionidae), are considered the primary
reservoir of merbecoviruses;\textsuperscript{3,17} but in the Americas,
where merbecoviruses are the only lineage present, they have only been
found in other bat taxa. At the coarsest scale, these heterogeneities
are lost, and betacoronavirus biogeography tracks the deep rifts in bat
evolutionary history---but within broad regions, the component
coevolutionary systems may have very different dynamics.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{20} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{21} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus},\textsuperscript{16} stressing
the need for spatially targeted monitoring and discovery.

@@

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{13} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{14} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{17} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{36} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{14,16} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{37}

@@

We downloaded the data on bats hosts of \emph{Betacoronavirus} from
\texttt{https://www.viralemergence.org/betacov} on
Apr.~2022,\textsuperscript{14} and filtered it to ``known'' hosts
(established before the emergence of SARS-CoV-2) and ``novel'' hosts
(confirmed through sampling and competence assays since the initial data
collection). The original database was assembled by a combination of
data mining and literature surveys, including automated alerts on the
``bats'' and ``coronavirus'' keywords to identify novel empirical
evidence of bats-betacoronaviruses associations; this yielded a total of
126 known hosts, 47 of which were novel hosts.

@@

\leavevmode\hypertarget{ref-Olival2017HosVir}{}%
\CSLLeftMargin{13. }
\CSLRightInline{Olival, K. J. \emph{et al.} Host and viral traits
predict zoonotic spillover from mammals. \emph{Nature} \textbf{546},
646--650 (2017).}

\leavevmode\hypertarget{ref-Becker2022OptPre}{}%
\CSLLeftMargin{14. }
\CSLRightInline{Becker, D. J. \emph{et al.} Optimising predictive models
to prioritise viral discovery in zoonotic reservoirs. \emph{The Lancet
Microbe} (2022)
doi:\href{https://doi.org/10.1016/S2666-5247(21)00245-7}{10.1016/S2666-5247(21)00245-7}.}

\leavevmode\hypertarget{ref-Allen2017GloHot}{}%
\CSLLeftMargin{15. }
\CSLRightInline{Allen, T. \emph{et al.} Global hotspots and correlates
of emerging zoonotic diseases. \emph{Nature Communications} \textbf{8},
(2017).}

\leavevmode\hypertarget{ref-Cohen2022SamStr}{}%
\CSLLeftMargin{16. }
\CSLRightInline{Cohen, L. E., Fagre, A. C., Chen, B., Carlson, C. J. \&
Becker, D. J. Sampling strategies and pre-pandemic surveillance gaps for
bat coronaviruses. 2022.06.15.496296 (2022)
doi:\href{https://doi.org/10.1101/2022.06.15.496296}{10.1101/2022.06.15.496296}.}

\leavevmode\hypertarget{ref-Latinne2020OriCro}{}%
\CSLLeftMargin{17. }
\CSLRightInline{Latinne, A. \emph{et al.} Origin and cross-species
transmission of bat coronaviruses in China. \emph{bioRxiv: The Preprint
Server for Biology} 2020.05.31.116061 (2020)
doi:\href{https://doi.org/10.1101/2020.05.31.116061}{10.1101/2020.05.31.116061}.}

>>>>

\hypertarget{bat-and-betacoronavirus-biogeography-are-broadly-consistent}{%
\subsection{Bat and betacoronavirus biogeography are broadly
consistent}\label{bat-and-betacoronavirus-biogeography-are-broadly-consistent}}

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{2,3} Projecting bat and betacoronavirus
phylogeny over space (fig.~\ref{fig:biogeo}), we find support for the
idea that bat community assembly is directly responsible for a global
mosaic of viral evolution. The distinct groupings (represented by
different colors, symbolizing positions in a subspace formed by the
first two phylogenetic principal components) are essentially equivalent
between the two groups, and can be coarsely delineated as (1) south and
southeast Asia, (2) east Asia (including northern China), west Asia, and
the Mediterranean coast; (3) Eurasia above a northing of 40; and (4)
Africa and south America. In some cases, this diverges from expectations
about coronavirus biogeography: for example, previous work has rarely
flagged India as a region of interest, but for both bats and
betacoronaviruses, the subcontinent falls into the same regions as the
southeast Asian peninsula (and indeed, the region is home to known bat
hosts of nobecoviruses, sarbecoviruses, and
merbecoviruses).\textsuperscript{3}

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are largely consistent. This may
be surprising, given that cospeciation plays a minor role in coronavirus
diversification,\textsuperscript{2} a property that would theoretically
allow for substantial broad divergence in their biogeography. However,
host jumps at the family level or higher are relatively rare and
significant events in coronavirus evolutionary
history;\textsuperscript{2,13} as a result, the mosaic of
betacoronavirus phylogeography is assembled from a set of overlapping
smaller coevolutionary systems, superimposed in space and filtered by
the importance of different subgroups in local host communities. For
example, the most speciose and cosmopolitan family of bats, the vesper
bats (Vespertilionidae), are considered the primary reservoir of
merbecoviruses;\textsuperscript{3,13} but in the Americas, where
merbecoviruses are the only lineage present, they have only been found
in other bat taxa. At the coarsest scale, these heterogeneities are
lost, and betacoronavirus biogeography tracks the deep rifts in bat
evolutionary history---but within broad regions, the component
coevolutionary systems may have very different dynamics.

@@

Bats, the second most diverse groups of mammals, are found worldwide;
gradients in their species richness generally track broader patterns of
mammal diversity, with a striking Neotropical hotspot (especially in the
Amazon basin) and a secondary hotspot centered in the southeast Asian
peninsula. These hotspots of bat diversity are generally presumed to be
hotspots of viral adaptive radiation, and therefore areas of concern for
human health.\textsuperscript{2,14} However, the hotspots of bat
betacoronavirus reservoirs show a distinct pattern, with primary
hotspots (both in terms of size and higher values) of host richness
situated primarily southeast Asia, parts of southern Europe, and to a
lesser extent parts of Africa in the -25-0 range of latitudes
(fig.~\ref{fig:richness}; top). Although hundreds of species likely host
undiscovered betacoronaviruses, machine learning predictions have
suggested that these undiscovered reservoirs should follow the same
diversity gradient.\textsuperscript{15} In principle, these hotspots of
locally-diverse, virus-rich bat communities should drive more adaptive
diversification in their viruses.

@@

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South
America has the most evolutionary distinct hosts \emph{and} viruses,
followed by secondary hotspots in southeast Asia and the Rift Valley
region have mostly distinct viruses. Some degree of sampling bias may
contribute to these patterns: for example, South-America is one of the
places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{2,14,16} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{17}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there is
obvious heterogeneity in betacoronavirus evolutionary distinctiveness
that is distinct from overall bat diversity.

Overall, these patterns recapitulate the evolutionary history of both
the order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe
bats (Rhinolophidae) are both the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}), the group of pandemic threats that have
been of the greatest interest to researchers\textsuperscript{13} (and so
have been sampled most intensively).\textsuperscript{17} The hotspots of
host richness and viral diversity in southeast Asia---both of which are
disproportionately high, considering the global landscape of bat species
richness---are almost entirely driven by viral adaptive radiation
through host switching within this clade\textsuperscript{3,15}. In
contrast, the Neotropical hotspot of viral distinctiveness is driven by
isolation by host vicariance. Out of the four main groups of
betacoronaviruses, only the subgenus \emph{Merbecovirus} (MERS-like
viruses) has been found in animals in the Americas---an introduction
that is generally presumed to be ancient.\textsuperscript{3} While
comparatively understudied, New World merbecoviruses have been found in
the ghost-faced bats (Mormoopidae), New World leaf-nosed bats
(Phyllostomidae), and free-tailed bats (Molossidae) (add cite: Olival
2020 PLoS Pathogens). The former two groups are endemic to the
Neotropics, while the explosive adaptive radiations of the latter two
(and particularly the phyllostomids) are responsible for the hotspot of
bat diversity in the Amazon. Together, these clades of New World bats
play host to a distinct regime of betacoronavirus coevolution.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{20} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{21} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus},\textsuperscript{17} stressing
the need for spatially targeted monitoring and discovery.

@@

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{14} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{15} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{13} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{36} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{15,17} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{37}

@@

We downloaded the data on bats hosts of \emph{Betacoronavirus} from
\texttt{https://www.viralemergence.org/betacov} on
Apr.~2022,\textsuperscript{15} and filtered it to ``known'' hosts
(established before the emergence of SARS-CoV-2) and ``novel'' hosts
(confirmed through sampling and competence assays since the initial data
collection). The original database was assembled by a combination of
data mining and literature surveys, including automated alerts on the
``bats'' and ``coronavirus'' keywords to identify novel empirical
evidence of bats-betacoronaviruses associations; this yielded a total of
126 known hosts, 47 of which were novel hosts.

@@

\leavevmode\hypertarget{ref-Latinne2020OriCro}{}%
\CSLLeftMargin{13. }
\CSLRightInline{Latinne, A. \emph{et al.} Origin and cross-species
transmission of bat coronaviruses in China. \emph{bioRxiv: The Preprint
Server for Biology} 2020.05.31.116061 (2020)
doi:\href{https://doi.org/10.1101/2020.05.31.116061}{10.1101/2020.05.31.116061}.}

\leavevmode\hypertarget{ref-Olival2017HosVir}{}%
\CSLLeftMargin{14. }
\CSLRightInline{Olival, K. J. \emph{et al.} Host and viral traits
predict zoonotic spillover from mammals. \emph{Nature} \textbf{546},
646--650 (2017).}

\leavevmode\hypertarget{ref-Becker2022OptPre}{}%
\CSLLeftMargin{15. }
\CSLRightInline{Becker, D. J. \emph{et al.} Optimising predictive models
to prioritise viral discovery in zoonotic reservoirs. \emph{The Lancet
Microbe} (2022)
doi:\href{https://doi.org/10.1016/S2666-5247(21)00245-7}{10.1016/S2666-5247(21)00245-7}.}

\leavevmode\hypertarget{ref-Allen2017GloHot}{}%
\CSLLeftMargin{16. }
\CSLRightInline{Allen, T. \emph{et al.} Global hotspots and correlates
of emerging zoonotic diseases. \emph{Nature Communications} \textbf{8},
(2017).}

\leavevmode\hypertarget{ref-Cohen2022SamStr}{}%
\CSLLeftMargin{17. }
\CSLRightInline{Cohen, L. E., Fagre, A. C., Chen, B., Carlson, C. J. \&
Becker, D. J. Sampling strategies and pre-pandemic surveillance gaps for
bat coronaviruses. 2022.06.15.496296 (2022)
doi:\href{https://doi.org/10.1101/2022.06.15.496296}{10.1101/2022.06.15.496296}.}
